---
title: "CUP Umaps"
author: "A.DeMartin"
date: "2025-03-17"
output:
  pdf_document:
    toc: true
  html_document:
    keep_md: true
    toc: true
    self_contained: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
options(width = 100)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, dev = c("png", "pdf"))
seed <- 1234
```



## load packages
```{r load packages, warning=FALSE, include=FALSE}
library(ExploreSCdataSeurat3)
library(runSeurat3)
library(Seurat)
library(ggpubr)
library(pheatmap)
library(SingleCellExperiment)
library(dplyr)
library(tidyverse)
library(viridis)
library(muscat)
library(circlize)
library(destiny)
library(scater)
library(metap)
library(multtest)
library(clusterProfiler)
library(org.Hs.eg.db)
library(msigdbr)
library(enrichplot)
library(DOSE)
library(grid)
library(gridExtra)
library(ggupset)
library(VennDiagram)
library(NCmisc)
```

## load file
```{r load merged file}
##load merged file 
fileNam <- "/R/CUP/data/CD45neg_allmerged_seurat.rds"
seuratM <- readRDS(fileNam)
table(seuratM$dataset)
table(seuratM$RNA_snn_res.0.25)
table(seuratM$orig.ident)

```


```{r add metadata}
### add any type of metadata
### patient
CUP_01 <- c("373291_01-1_20241211_Hu_CUP_01_CD45neg", "380131_01-1_20250224_Hu_CUP_01_stroma")
CUP_03 <- c("373291_07-7_20241211_Hu_CUP_03_CD45neg")
CUP_04 <- c("373291_03-3_20241211_Hu_CUP_04_CD45neg")
CUP_05 <- c("373291_05-5_20241211_Hu_CUP_05_CD45neg")
hLN_01 <- c("13_20200723_Hu_LN_Lin_neg_FSC_v3")
hLN_02 <- c("1_20200909_Human__LN_FSC1_v3","2_20200909_Human__LN_FSC2_v3")
hLN_03 <- c("1_20210113_Human_LN_FSC1_v3","2_20210113_Human_LN_FSC2_v3")
hLN_04 <- c("o27533_1_01-1_20220201_Hu_LN_stroma_v3")
hLN_05 <- c("o27533_1_05-5_20220202_Hu_LN_stroma_v3")

seuratM$patient <- "pat_nr"
seuratM$patient[which(seuratM$dataset %in% CUP_01)] <- "CUP_01"
seuratM$patient[which(seuratM$dataset %in% CUP_03)] <- "CUP_03"
seuratM$patient[which(seuratM$dataset %in% CUP_04)] <- "CUP_04"
seuratM$patient[which(seuratM$dataset %in% CUP_05)] <- "CUP_05"
seuratM$patient[which(seuratM$dataset %in% hLN_01)] <- "hLN_01"
seuratM$patient[which(seuratM$dataset %in% hLN_02)] <- "hLN-02"
seuratM$patient[which(seuratM$dataset %in% hLN_03)] <- "hLN_03"
seuratM$patient[which(seuratM$dataset %in% hLN_04)] <- "hLN_04"
seuratM$patient[which(seuratM$dataset %in% hLN_05)] <- "hLN_05"
table(seuratM$patient)

###order patients
ordpatients <- c("hLN_01","hLN-02", "hLN_03","hLN_04", "hLN_05","CUP_01", "CUP_03", "CUP_04","CUP_05")
seuratM$patient <- factor(seuratM$patient, levels=c("hLN_01","hLN-02", "hLN_03","hLN_04", "hLN_05","CUP_01", "CUP_03", "CUP_04","CUP_05"))
Idents(seuratM) <- seuratM$patient
table(seuratM$patient)


#### diseaseCond
seuratM$diseaseCond <- "diseaseCond"
seuratM$diseaseCond[grepl("CUP", seuratM$dataset)] <- "CUP"
seuratM$diseaseCond[grepl("LN", seuratM$dataset)] <- "healthy"
table(seuratM$diseaseCond)

###order
orddiseaseCond <- c("healthy", "CUP")
seuratM$diseaseCond <- factor(seuratM$diseaseCond, levels=c("healthy", "CUP"))
Idents(seuratM) <- seuratM$diseaseCond
table(seuratM$diseaseCond)




```


```{r data trimming}

## Remove outlier and filter out CD45+ clusters 

seuratMfil <- subset(seuratM, patient == c("hLN_01"), invert = TRUE) 

seuratMfil <- subset(seuratMfil, RNA_snn_res.0.25 %in% c("7","9","11","19"),invert = TRUE)

table(seuratMfil$patient)

Idents(seuratMfil) <- seuratMfil$RNA_snn_res.0.25
DimPlot(seuratMfil, reduction = "umap", pt.size = 0.1) 


```

```{r run UMAP total}

##rerun seurat
seuratMfil <- NormalizeData (object = seuratMfil)
seuratMfil <- FindVariableFeatures(object = seuratMfil)
seuratMfil <- ScaleData(object = seuratMfil, verbose = TRUE)
seuratMfil <- RunPCA(object=seuratMfil, npcs = 30, verbose = FALSE)
seuratMfil <- RunTSNE(object=seuratMfil, reduction="pca", dims = 1:20)
seuratMfil <- RunUMAP(object=seuratMfil, reduction="pca", dims = 1:20)
seuratMfil <- FindNeighbors(object = seuratMfil, reduction = "pca", dims= 1:20)

res <- c(0.25, 0.6, 0.8, 0.4)
for (i in 1:length(res)) {
  seuratMfil <- FindClusters(object = seuratMfil, resolution = res[i], random.seed = 1234)
}

```

### CUP non-hematopoietic cells {.tabset}
#### Resolution 0.25
```{r run UMAP plots_resolution}
Idents(seuratMfil) <- seuratMfil$RNA_snn_res.0.25
DimPlot(seuratMfil, reduction = "umap", pt.size = 0.1) 
```

#### patient
```{r run UMAP plots_patient}
Idents(seuratMfil) <- seuratMfil$patient
DimPlot(seuratMfil, reduction = "umap", pt.size = 0.1, shuffle = TRUE) 
```

#### diseaseCond
```{r run UMAP plots_diseaseCond}
Idents(seuratMfil) <- seuratMfil$diseaseCond
coldiseaseCond <- c("#202547","lightgrey")
names(coldiseaseCond) <- c("healthy", "CUP")
DimPlot(seuratMfil, reduction = "umap", pt.size = 0.1, cols = coldiseaseCond, order = "healthy") 

```


### plot features cluster marker 
```{r feature plot marker genes list}

genelist <- c("ENSG00000261371.PECAM1","ENSG00000117707.PROX1","ENSG00000163359.COL6A3","ENSG00000107796.ACTA2","ENSG00000164692.COL1A2","ENSG00000205420.KRT6A","ENSG00000119888.EPCAM","ENSG00000171345.KRT19","ENSG00000147889.CDKN2A")

```

### Marker genes from distinct cell types {.tabset}
#### Endothelial cells
```{r Endothelial cells}
FeaturePlot(seuratMfil, reduction = "umap", 
          features = genelist[1:2],
          cols=c("lightgrey", "darkred"),
          order = T )+
theme(legend.position="right", legend.title=element_text(size=3)) 
```

#### fibroblasts
```{r fibroblasts}
FeaturePlot(seuratMfil, reduction = "umap", 
          features = genelist[3:5],
          cols=c("lightgrey", "darkred"),
          order = T )+
theme(legend.position="right", legend.title=element_text(size=3)) 
```

#### Tumor cells
```{r Tumor cells}
FeaturePlot(seuratMfil, reduction = "umap", 
          features = genelist[6:9],
          cols=c("lightgrey", "darkred"),
          order = T )+
theme(legend.position="right", legend.title=element_text(size=3)) 
```

```{r feature plots total, eval=FALSE, include=TRUE}
##Endo
FeaturePlot(seuratMfil, features = "ENSG00000261371.PECAM1", pt.size = 0.5, cols = c("lightgrey", "#BE3144"))
FeaturePlot(seuratMfil, features = "ENSG00000117707.PROX1", pt.size = 0.5, cols = c("lightgrey", "#BE3144"))

##Fibroblastic
FeaturePlot(seuratMfil, features = "ENSG00000163359.COL6A3", pt.size = 0.5, cols = c("lightgrey", "#BE3144"))
FeaturePlot(seuratMfil, features = "ENSG00000107796.ACTA2", pt.size = 0.5, cols = c("lightgrey", "#BE3144"))
FeaturePlot(seuratMfil, features = "ENSG00000164692.COL1A2", pt.size = 0.5, cols = c("lightgrey", "#BE3144"))


##Tumor
FeaturePlot(seuratMfil, features = "ENSG00000205420.KRT6A", pt.size = 0.5, cols = c("lightgrey", "#BE3144"))
FeaturePlot(seuratMfil, features = "ENSG00000119888.EPCAM", pt.size = 0.5, cols = c("lightgrey", "#BE3144"))
FeaturePlot(seuratMfil, features = "ENSG00000171345.KRT19", pt.size = 0.5, cols = c("lightgrey", "#BE3144"))
FeaturePlot(seuratMfil, features = "ENSG00000147889.CDKN2A", pt.size = 0.5, cols = c("lightgrey", "#BE3144"))



```

### Assign cell types to clusters
```{r Assign cell types to clusters}

colclusterName <- c("#67001f", "#D53E4F", "#f4a582", "#FEE08B", "#003c30","#01665e","#66C2A5", "#3288BD","#BEAED4", "#c7eae5","#355C7D","#202547","#B45B5C")
names(colclusterName) <- c("FRC1","FRC2","Tu1","BEC1","CAF","BEC2", "PRC","PRC_Tu","BEC3","Tu2","Tu3","LEC","PI16_FRC")


##assign celltypes
seuratMfil$colclusterName <- "celltype"
seuratMfil$colclusterName[which(seuratMfil$RNA_snn_res.0.25 == "0")] <- "FRC1"
seuratMfil$colclusterName[which(seuratMfil$RNA_snn_res.0.25 == "1")] <- "FRC2"
seuratMfil$colclusterName[which(seuratMfil$RNA_snn_res.0.25 == "2")] <- "Tu1"
seuratMfil$colclusterName[which(seuratMfil$RNA_snn_res.0.25 == "3")] <- "BEC1"
seuratMfil$colclusterName[which(seuratMfil$RNA_snn_res.0.25 == "4")] <- "CAF"
seuratMfil$colclusterName[which(seuratMfil$RNA_snn_res.0.25 == "5")] <- "BEC2"
seuratMfil$colclusterName[which(seuratMfil$RNA_snn_res.0.25 == "6")] <- "PRC"
seuratMfil$colclusterName[which(seuratMfil$RNA_snn_res.0.25 == "7")] <- "PRC_Tu"
seuratMfil$colclusterName[which(seuratMfil$RNA_snn_res.0.25 == "8")] <- "BEC3"
seuratMfil$colclusterName[which(seuratMfil$RNA_snn_res.0.25 == "9")] <- "Tu2"
seuratMfil$colclusterName[which(seuratMfil$RNA_snn_res.0.25 == "10")] <- "Tu3"
seuratMfil$colclusterName[which(seuratMfil$RNA_snn_res.0.25 == "11")] <- "LEC"
seuratMfil$colclusterName[which(seuratMfil$RNA_snn_res.0.25 == "12")] <- "PI16_FRC"

table(seuratMfil$colclusterName)
```


### UMAP with defined cluster name {.tabset}
#### UMAP with clusterName
```{r UMAP with clusterName}
Idents(seuratMfil) <- seuratMfil$colclusterName
DimPlot(seuratMfil, reduction = "umap", pt.size = 0.1, cols = colclusterName, label =TRUE) 
```

#### UMAP with clusterName split by disease
```{r UMAP with clusterName split by disease}
Idents(seuratMfil) <- seuratMfil$colclusterName
DimPlot(seuratMfil, reduction = "umap", pt.size = 0.1, cols = colclusterName, label =FALSE, split.by = "diseaseCond") 
```

#### UMAP with clusterName split by patients
```{r UMAP with clusterName split by patients, fig.width= 12}
Idents(seuratMfil) <- seuratMfil$colclusterName
DimPlot(seuratMfil, reduction = "umap", pt.size = 0.1, cols = colclusterName, label =FALSE, split.by = "patient") 
```

### calculate cluster marker genes
```{r marker genes_total, include=TRUE, eval=FALSE}
##cluster marker
Idents(seuratMfil) <- seuratMfil$colclusterName
markerGenes <- FindAllMarkers(seuratMfil, only.pos=T) %>% 
  dplyr::filter(p_val_adj < 0.01)
#save table
write.table(markerGenes, 
            file= "/R/CUP/analysis/markerGenes",
            sep="\t",
            quote=F,
            row.names=F,
            col.names=T)
```


### Plot gene signatures onto UMAP
```{r cell type gene signatures}

##convert seurat object to sce object
sce <- as.SingleCellExperiment(seuratMfil)


signatures <- list(
  CAF = c("PDPN", "POSTN", "FAP", "LRRC15"),
  FRC = c("C7", "CCL21","CCL19", "CXCL12","IL33", "PI16"),
  PRC = c("ACTA2", "ITGA7","RGS5", "NOTCH3", "MCAM"),
  Tumor = c("KRT6A","EPCAM","KRT1","CDKN2A"),
  Endothelial =c("PECAM1", "VWF")
)


# Extract gene symbols
genes <- data.frame(geneID = rownames(sce)) %>%
  mutate(gene = gsub(".*\\.", "", geneID))

# Function to compute signature and return UMAP plot
plot_signature_umap <- function(sig_name, sig_genes) {
  signGenes <- genes %>% filter(gene %in% sig_genes)
  sceSub <- sce[rownames(sce) %in% signGenes$geneID, ]
  
  cntMat <- rowSums(t(as.matrix(sceSub@assays@data$logcounts))) / nrow(signGenes)
  sceSub$sign <- cntMat
  sceSub$sign2 <- pmin(sceSub$sign, 2.5)  # cap at 2.5
  
  plotUMAP(sceSub, colour_by = "sign2", point_size = 1) +
    scale_color_viridis_c() +  # or another color scale
    ggtitle(paste0(sig_name, " signature")) +
    theme_minimal() +
    theme(legend.position = "right")
}

# Generate all plots
umap_plots <- lapply(names(signatures), function(name) {
  plot_signature_umap(name, signatures[[name]])
})

```

### Gene signatures of cell types {.tabset}
#### CAF gene signatures
```{r CAF gene signatures}

print(umap_plots[[1]])

```

#### FRC gene signatures
```{r FRC gene signatures}

print(umap_plots[[2]])

```

#### PRC gene signatures
```{r PRC gene signatures}

print(umap_plots[[3]])

```

#### Tumor gene signatures
```{r Tumor gene signatures}

print(umap_plots[[4]])

```

#### Endothelial gene signatures
```{r Endothelial gene signatures}

print(umap_plots[[5]])

```

### dotplot marker genes
```{r dotplot marker genes_total, fig.height=8, fig.width=8}
seuratMfil$colclusterName <- factor(seuratMfil$colclusterName, levels=c("PI16_FRC","FRC1","FRC2", "PRC","PRC_Tu","CAF","Tu1","Tu2","Tu3","BEC1","BEC2","BEC3","LEC"))
Idents(seuratMfil) <-seuratMfil$colclusterName

genes <- data.frame(gene=rownames(seuratMfil)) %>% 
  mutate(geneID=gsub("^.*\\.", "", gene))

selGenes <- data.frame(geneID=rev(c( "COL1A2", "COL6A3", "PDGFRB", "DCN",  "FBN1", "C7", "CCL21","CCL19", "CXCL12","IL33", "PI16", "ACTA2", "ITGA7","RGS5", "NOTCH3", "MCAM","PDPN", "FAP","LRRC15","POSTN","PECAM1", "VWF", "PROX1"))) %>% left_join(., genes, by="geneID") %>% filter(gene != "ENSG00000232995.RGS5") 


DotPlot(seuratMfil, features = selGenes$gene, group.by = "colclusterName") +
  RotatedAxis() +
  scale_color_viridis(option = "F") +
  coord_flip()



```

### relative abundance of cell types 
```{r rel abundance_total, fig.height=8, fig.width=8}
###dataset
datList <- NULL
for(con in unique(seuratMfil$patient)){
  seuratSub <- subset(seuratMfil, patient==con)
  print(dim(seuratSub))
  dat_con <- as.data.frame(table(seuratSub$colclusterName)) %>%
  mutate(percent=Freq/ncol(seuratSub)) %>% mutate(patient=con)
  datList[[con]] <- dat_con
}

dat_all <- do.call("rbind", datList)


# Define the order you want
patient_order <- c("hLN_01", "hLN-02", "hLN_03", "hLN_04", "hLN_05", "CUP_01", "CUP_03", "CUP_04", "CUP_05")  # customize as needed
dat_all$patient <- factor(dat_all$patient, levels = patient_order)


## plot abundance - separated by patients
ggbarplot(dat_all, x= "patient", y= "percent", fill = "Var1",  legend = "right", legend.titel = "cluster", ylab = "frequency")  + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))



###cond
###patient_diseaseCond
datList <- NULL
for(con in unique(seuratMfil$diseaseCond)){
  seuratSub <- subset(seuratMfil, diseaseCond==con)
  print(dim(seuratSub))
  dat_con <- as.data.frame(table(seuratSub$colclusterName)) %>%
  mutate(percent=Freq/ncol(seuratSub)) %>% mutate(diseaseCond=con)
  datList[[con]] <- dat_con
}

dat_all <- do.call("rbind", datList)


## plot abundance - separated by condition
ggbarplot(dat_all, x= "diseaseCond", y= "percent", fill = "Var1", palette = colclusterName, legend = "right", legend.titel = "cluster", ylab = "frequency")  + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 



```

### Filter the object with fibroblasts only
```{r fibroblasts only}

## Filter particular cell populations 
seuratM_fib_only <- subset(seuratMfil, colclusterName %in% c("FRC1","FRC2","PRC","PI16_FRC","CAF","PRC_Tu")) 

colname_Fib <- c("#67001f", "#D53E4F", "#003c30", "#3288BD","#BEAED4", "#c7eae5")
names(colname_Fib) <- c("FRC1","FRC2","CAF","PRC","PRC_Tu","PI16_FRC")

##rerun seurat
seuratM_fib_only <- NormalizeData (object = seuratM_fib_only)
seuratM_fib_only <- FindVariableFeatures(object = seuratM_fib_only)
seuratM_fib_only <- ScaleData(object = seuratM_fib_only, verbose = TRUE)
seuratM_fib_only <- RunPCA(object=seuratM_fib_only, npcs = 30, verbose = FALSE)
seuratM_fib_only <- RunTSNE(object=seuratM_fib_only, reduction="pca", dims = 1:20)
seuratM_fib_only <- RunUMAP(object=seuratM_fib_only, reduction="pca", dims = 1:20)
seuratM_fib_only <- FindNeighbors(object = seuratM_fib_only, reduction = "pca", dims= 1:20)

res <- c(0.25, 0.6, 0.8, 0.4)
for (i in 1:length(res)) {
  seuratM_fib_only <- FindClusters(object = seuratM_fib_only, resolution = res[i], random.seed = 1234)
}

Idents(seuratM_fib_only) <- seuratM_fib_only$colclusterName
DimPlot(seuratM_fib_only, reduction = "umap", pt.size = 0.1) 

Idents(seuratM_fib_only) <- seuratM_fib_only$patient
DimPlot(seuratM_fib_only, reduction = "umap", pt.size = 0.1, shuffle = TRUE) 

Idents(seuratM_fib_only) <- seuratM_fib_only$diseaseCond
coldiseaseCond <- c("#202547","lightgrey")
names(coldiseaseCond) <- c( "CUP", "healthy")
DimPlot(seuratM_fib_only, reduction = "umap", pt.size = 0.1, cols = coldiseaseCond, order = "CUP") 
```

### relative abundance of cell types {.tabset} 
#### rel abundance_fibroblasts spilt by patients 
```{r rel abundance_fibroblasts spilt by patients, fig.height=8, fig.width=8}
datList <- NULL
for(con in unique(seuratM_fib_only$patient)){
  seuratSub <- subset(seuratM_fib_only, patient==con)
  print(dim(seuratSub))
  dat_con <- as.data.frame(table(seuratSub$colclusterName)) %>%
  mutate(percent=Freq/ncol(seuratSub)) %>% mutate(patient=con)
  datList[[con]] <- dat_con
}

dat_all <- do.call("rbind", datList)

# Define the order you want
patient_order <- c("hLN_01", "hLN-02", "hLN_03", "hLN_04", "hLN_05", "CUP_01", "CUP_03", "CUP_04", "CUP_05")  # customize as needed
dat_all$patient <- factor(dat_all$patient, levels = patient_order)

# plot abundance - separated by patients
ggbarplot(dat_all, x= "patient", y= "percent", fill = "Var1", palette = colname_Fib,  legend = "right", legend.titel = "cluster", ylab = "frequency")  + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

#### rel abundance_fibroblasts spilt by condition 
```{r rel abundance_fibroblasts spilt by condition, fig.height=8, fig.width=8}
datList <- NULL
for(con in unique(seuratM_fib_only$diseaseCond)){
  seuratSub <- subset(seuratM_fib_only, diseaseCond==con)
  print(dim(seuratSub))
  dat_con <- as.data.frame(table(seuratSub$colclusterName)) %>%
  mutate(percent=Freq/ncol(seuratSub)) %>% mutate(diseaseCond=con)
  datList[[con]] <- dat_con
}

dat_all <- do.call("rbind", datList)

# plot abundance - separated by condition
ggbarplot(dat_all, x= "diseaseCond", y= "percent", fill = "Var1", palette = colname_Fib, legend = "right", legend.titel = "cluster", ylab = "frequency")  + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

## session info
```{r date and session info}
date()
sessionInfo()
```
